Comparing summed probability distributions of shoreline and radiocarbon dates from the Stone Age on the Norwegian Skagerrak coast

### 2023-04-21 analysis/paper/paper.Rmd
ADMUR for SSPD

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(kableExtra)
library(sf)
```

```{r data, include = FALSE}
surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
```

# Introduction
Population size is regarded as one of the primary drivers of cultural variation, and is of critical importance to our understanding of past human societies [@shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and beings to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

# Background
Large parts of the post-glacial landscape of Northern Scandinavia is characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. As coastal foragers appear to have predominantly settled on or close to the contemporaneous shoreline, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement---a method known as shoreline dating. This is not least useful for dating the large number of surveyed sites in the region where other temporal data that often follow with an excavation are not available, such as radiocarbon dates or typological indicators in artefact inventories.

The frequency of shoreline dated sites has also been compared to RSPDs in the past [@solheim2018; @tallavaara2020]. However, this has been done by finding point estimates of shoreline dates that are then aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not take into account the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility. Thus, the geographical and temporal limits outlined above are also used for this study.

@@

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represents the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data---the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

@@

The procedure outlined above represent the standard approach when modelling population dynamics in archaeology. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally testing the best model using the Monte Carlo approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these. One major benefit of the approach is thus that it provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet known to be older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites. This gave a total of  shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates. Following both from the point made by @freeman concerning the comparison of radiocarbon dates from populations with different , as noted above,  and from the fact that shoreline dating appears less reliable after 4000 BCE and looses its utility c. 2500 BCE, the Mesolithic was the main focus of this analysis. Consequently, calibrated ^14^C-dates with a probability falling later than 2500 BCE were excluded from the analysis. Dates falling after 4000 BCE, and the first small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects, but these results should be treated with for the reasons noted above.

@@

```{r map, echo = FALSE, fig.cap = "Map of the study area and analysed sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1. Of the excavated sites, 102 have been dated by means of shoreline dating.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/map.png"))
```

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other. When calibrated, these are summed and divided by the number of dates in the bin before being included in the final RSPD. Furthermore, the ^14^C-dates were not normalised. This follows from a peculiarity in the calibration procedure which can lead the probability mass of a date to sum to more than one [@weninger2015]. As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD [e.g. @crema2021]. All calibration was done using the IntCal20 calibration curve [@reimer2020].

The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by maximum likelihood search, using the differential evolution optimization algorithm DEoptimR. For each model, a series of individual calendar years are drawn from the distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 1000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate). Four geologic reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here. These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the sea-level to the location of a site to be dated based on it distance from the isobases of the displacement curves. The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated by a gamma distribution with shape =  and scale = . When shoreline dating, the gamma distribution is here sequentially stepped through at increments of 0.1m and transferred to the calendar year scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure gives the the shoreline date of a site [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1m and the calendar scale is kept at a resolution of 5 years.

```{r shoredate, echo = FALSE, fig.cap = "A) Example location relative to the isobases of the displacement curves.  B) The geologic displacement curves and the curve interpolated to the example location. C) Resulting shoreline date in light grey on the x-axis. The black line underneath marks the 95\\% HDR. The dashed line marks the elevation of the example location. The exponential function on the y-axis decays with ratio $\\lambda$ and represents the likely elevation of a site above sea-level when it was in use. ", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

In Figure  the shoreline dated sites are plotted according to their elevation. However, due to the variable uplift rates within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution. As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east. If the elevation of the sites is instead transformed to point-estimated shoreline dates, done by coupling the mean elevation of the site and the mean of the displacement curve interpolated to each site location, the temporal distribution instead looks like the one provided in figure. This follows the lines of how shoreline dates have usually been treated when the frequency distribution of shoreline dated sites has been investigated in the past. However, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates. This problem is further accentuated by the fact that shoreline dates are associated with more uncertainty than ^14^C-dates. Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east, this means that this also results in both a geographical and temporal bias in the final results. The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure.

In much the same way as characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites (Figure \@ref(fig:incpolys)). These intervals were then assigned a weight based on how the density of observed sites is distributed among them.

@@

Following from the commutative nature of the shoreline dates, the Monte Carlo simulation is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution of the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites these were then summed, and the entire process repeated a total of 1000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

## Continous piece-wise linear models
The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of theoretical models that , and was an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics, published with the accompanying  *ADMUR* R package. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.

@@

```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model. B) Same as A, but with a exponential model. C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model. D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
Comparing the the directions and devitions from the standard models, fig indicates some process of overall general increase in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time. However, with global p-values all below X,  all models can be rejected as explaining the data they have been fit to. Based on common demographic developments and previous suggestions, there is some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments. However, as introduced above, several authors have warned against over-interpreting such deviation. Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.

@@

>>>>

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(kableExtra)
library(sf)
library(shoredate)
```

```{r data, include = FALSE}
surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
```

# Introduction
Population size is regarded as one of the primary drivers of cultural variation, and is of critical importance to our understanding of past human societies [@shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018a; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021, see also @persson2009]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

# Background
Large parts of the post-glacial landscape of Northern Scandinavia is characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. As coastal foragers appear to have predominantly settled on or close to the contemporaneous shoreline, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. This is not least useful for dating the large number of surveyed sites in the region where other temporal data that often follow with an excavation are not available, such as radiocarbon dates or typological indicators in artefact inventories.

The frequency of shoreline dated sites has also been compared to RSPDs in the past [@solheim2018; @tallavaara2020]. However, this has been done by finding point estimates of shoreline dates that are then aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not take into account the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline, followed by a clear break around 2500 BCE, at which point shoreline dating appears to loose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represents the combination of events and uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that an SPD cannot be directly analysed to draw inferences on population dynamics, nor can it be directly compared to other time-series data [@carleton2018; @timpson2021]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

@@

The procedure outlined above represent the standard approach when modelling population dynamics in archaeology. This can be helpful for determining whether the population behaviour has followed the trajectories of standard theoretical models, and is a useful first step here as few assumptions concerning the behaviour and overall development of number of sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics based RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure of model comparison, and finally testing the best model using the Monte Carlo approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of the population proxy changes direction, thus allowing for the identification of key events with a conservative description of population dynamics between these. The approach thus also provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were therefore subsequently subjected to the approach presented by @timpson2021.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2022. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites. This gave a total of  shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates. Following both from the point made by @freeman concerning the comparison of radiocarbon dates from populations with different , as noted above,  and from the fact that shoreline dating appears less reliable after 4000 BCE and looses its utility c. 2500 BCE, the Mesolithic was the main focus of this analysis. Consequently, calibrated ^14^C-dates with a probability falling later than 2500 BCE were excluded from the analysis. Dates falling after 4000 BCE and the first small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

@@

```{r map, echo = FALSE, fig.cap = "Map of the study area and shoreline dated sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1.", out.width = "500px", fig.align="center", regfloat = TRUE}

# Map prepared in 01_shoreline_spd.R 
knitr::include_graphics(here("analysis/figures/map.png"))
```

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the radiocarbon dates were initially binned using the *binPrep()* function from *rcarbon*, where dates were aggregated on a site by site basis if they fell within 200 uncalibrated ^14^C years of each other. When calibrated, these are summed and divided by the number of dates in the bin before being included in the final RSPD. Furthermore, the ^14^C-dates were not normalised. This follows from a peculiarity in the calibration procedure which can lead the probability mass of a date to sum to more than one [@weninger2015]. As a consequence of this, normalising the dates before summing has been shown to create spurious spikes in the RSPD [e.g. @crema2021]. All calibration was done using the IntCal20 calibration curve [@reimer2020].

The RSPD was first subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*. The Monte Carlo simulations were then performed using the *modelTest()* function from *rcarbon*. For each model, a series of individual calendar years are drawn from the distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate). Four geologic reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here. These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the location to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the sea-level to the location of a site to be dated based on it distance from the isobases of the displacement curves. The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE were found to have a likely elevation above sea-level that was reasonably approximated by a gamma distribution with shape =  and scale = . When shoreline dating, the gamma distribution is here sequentially stepped through at increments of 0.1m and transferred to the calendar year scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure gives the the shoreline date of a site [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

```{r shoredate, echo = FALSE, fig.cap = "A) Example location relative to the isobases of the displacement curves.  B) The geologic displacement curves and the curve interpolated to the example location. C) Resulting shoreline date in light grey on the x-axis. The black line underneath marks the 95\\% HDR. The dashed line marks the elevation of the example location. The exponential function on the y-axis decays with ratio $\\lambda$ and represents the likely elevation of a site above sea-level when it was in use. ", out.width = "500px", fig.align="center", regfloat = TRUE}
target_site <- sites[1,]
target_plot(target_site)

```

@@

In Figure  the shoreline dated sites are plotted according to their elevation. However, due to the variable uplift rates within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly tell us much about their temporal distribution. As the trajectory for sea-level regression is starker towards the north-east, the same elevation at a location to the south-west implies a younger date that one towards the north-east. If the elevation of the sites is instead transformed to point-estimated shoreline dates by finding the mean shoreline date for the sites, without accounting for the likely distance to the shoreline, the temporal distribution instead looks like the one provided in Figure. This is similar to how shoreline dates have been treated when the frequency distribution of shoreline dated sites has been investigated in the past [@breivik; @solheim; @fossum; @mjærum; @tallavaara]. However, apart from not accounting for the distance between the site and the sea, this does not properly account for the uncertainty associated with the dates, and is thus similar to summarising point-estimates of ^14^C-dates. This problem is further accentuated by the fact that shoreline dating, as implemented here, is associated with more uncertainty than ^14^C-dates. Additionally, given that the displacement curves have a more rapid drop early in the Holocene and at locations further to the north-east in the study area, this means that this also results in both a geographical and temporal bias in the final results. The SSPD resulting from using the probabilistic method of shoreline dating is given in Figure, which clearly demonstrates that accounting for these dimensions gives a different result compared to the traditional approach.

However, in the same way is with RSPDs, the SSPD cannot be directly interpreted as reflecting the intensity of sites over time, due to the summation issues introduced above. Furthermore, similar to how characteristics of the calibration curve can introduce bias to the RSPD, the same is true for the SSPD, where the local trajectory of relative sea-level change effectively functions as the calibration curve for each site. As the shoreline displacement curves are interpolated to the sites based on their location along a south-west--north-east gradient, each site is effectively associated with a unique shoreline displacement curve, provided they are not located on exactly the same isobase. With analogy to the radiocarbon methodology, this would be equivalent to each date being associated with a unique calibration curve. As it would be computationally prohibitive to interpolate the shoreline displacement trajectory for each date to be simulated in the Monte Carlo procedure, one shoreline displacement curve was initially interpolated to the centre of each of a series of 2 km wide line segments running perpendicular to the shoreline gradient between the extremes of the distribution of sites. These intervals were then assigned a weight based on how the density of observed sites is distributed among them (Figure \@ref(fig:incpolys)).

@@

Following from the commutative nature of the shoreline dates, the Monte Carlo simulation is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any individual year is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

## Continous piece-wise linear models
The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of theoretical models that , and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.

@@

```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model. B) Same as A, but with a exponential model. C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model. D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
Comparing the the directions and devitions from the three standard models, Figure appears to indicate some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time. However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD. Based on common demographic developments and previous suggestions, there is possibly some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments. However, as introduced above, several authors have warned against over-interpreting such deviation. Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.

@@

@berg-hansen2022 have argued that there is an increase in sites in the first half of the MM in the northern part of the study area of this paper. However, their analysis is seemingly also based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. This analysis instead indicates a drop in the number of sites through this time-interval
